A Sound Deep Embedding of Arbitrary Normal Modal Logics in Coq

### 2022-09-04 paper/paper.tex
Rename s/form/formula/ on the paper

\paragraph{Syntax and semantics of modal logics}
To implement our library, we use a \emph{deep embedding} of
formulas, i.e. we represent formulas as the following Coq inductive datatype.
\begin{minted}{coq}
Inductive form: Set :=
| Lit    : nat          -> form
| Neg    : form -> form
| Box    : form -> form
| Dia    : form -> form
| And    : form -> form -> form
| Or     : form -> form -> form
| Implies: form -> form -> form.
\end{minted}
Most constructors of the type have an immediate meaning, since they represent
connectives of propositional logics. The constructor \mintinline{coq}{Lit}
is used to construct propositional variables; constructor \mintinline{coq}{Box}
denotes the modal operator $\Box$ and \mintinline{coq}{Dia}, the operator
$\Diamond$. Furthermore, the type \mintinline{coq}{theory} is defined as a list
of formulas.

\begin{minted}{coq}
Definition theory := list form.
\end{minted}

Before defining the Kripke semantics, we need to represent the model used to
interpret a modal logic formula. We encode a frame by a record formed by
a type \mintinline{coq}{W: Set} to represent the set of worlds and an
accessibility relation denoted by a type \mintinline{coq}{R: W -> W -> Prop}.
\begin{minted}{coq}
Record Frame: Type := {
  W: Set;
  R: W -> W -> Prop;
}.
\end{minted}
A model is then denoted by another record, formed by a frame and a labeling function,
which assign the truth value of a variable at a given world.
\begin{minted}{coq}
Record Model: Type := {
  F: Frame; 
  v: nat -> (W F) -> Prop; 
}.
\end{minted}
Note that the type of the labeling function \mintinline{coq}{v} ensures that it
refers to a world from the set of possible worlds in the frame
\mintinline{coq}{F}\footnote{We recall that \mintinline{coq}{W F} is Coq's syntax for
accessing field \mintinline{coq}{W} from the record \mintinline{coq}{F}.}. The encoding of the Kripke semantics is done by a Coq function
that implements the relation of Figure~\ref{fig:kripke}. We omit the code
for propositional logics connectives for brevity. The interesting parts are for
variables (constructor \mintinline{coq}{Lit}) which uses the labeling function
\mintinline{coq}{v} from the model \mintinline{coq}{M}. The meaning of
modalities $\Box$ and $\Diamond$ are represented by a quantifying over worlds
using the accessibility relation $R$. 
\begin{minted}{coq}
Fixpoint formula_valuation (M: Model) (w: W (F M))
                                 (phi: form): Prop :=
  match phi with
  | Lit     x
     => v M x w (* should be inhabited, i.e., true *)
  | Box     psi
     => forall w': W (F M), R (F M) w w' ->
                            formula_valuation M w' psi
  | Dia     psi
     => exists w': W (F M), R (F M) w w' /\
                            formula_valuation M w' psi
  (* some code omitted *)
  end.
\end{minted}
In order to ease the task of writing formulas and its semantics properties, we
define Coq notations that mimics the standard way of writing them.

\paragraph{The Hilbert deductive system}
Basically, a Hilbert-style deductive system allows the use of four
possibilities in a proof: 1) the use of a hypothesis; 2) the use of an axiom;
3) the use of the modus ponens rule or 4) the use of the generalization (necessity)
rule. The inductive type \mintinline{coq}{deduction} encodes these possibilities
as its constructors. The \mintinline{coq}{deduction} type is parameterized by
the set of axioms that form a system for some modal logic. In this way, our definition can be
used by several systems alike.
\begin{minted}{coq}
Inductive deduction (A: axiom -> Prop)
                           : theory -> form -> Prop :=
  (* Premise. *)
  | Prem: forall (t: theory)
                 (f: form)
                 (i: nat),
          (nth_error t i = Some f) -> deduction A t f
  (* Axiom. *)
  | Ax: forall (t: theory)
               (a: axiom)
               (f: form),
        A a -> instantiate a = f -> deduction A t f
  (* Modus Ponens. *)
  | Mp: forall (t: theory)
               (f g: form)
               (d1: deduction A t (f -> g))
               (d2: deduction A t f),
        deduction A t g
  (* Generalization. *)
  | Nec: forall (t: theory)
                (f: form)
                (d1: deduction A t f),
         deduction A t (Box f).
\end{minted}
The constructors \mintinline{coq}{Prem}, \mintinline{coq}{Mp} and \mintinline{coq}{Nec}
represent rules for hypothesis, modus ponens and generalization, respectively.
Before discussing the \mintinline{coq}{Ax}, we need to discuss how axioms are
represented in our library. The inductive type \mintinline{coq}{axiom} specifies
the different axioms supported by our library.
\begin{minted}{coq}
Inductive axiom: Set :=
  | ax1  : form -> form -> axiom
  | ax2  : form -> form -> form  -> axiom
  (* some code omitted *)
  | axK  : form -> form -> axiom
  | axPos: form -> form -> axiom
  | axT  : form -> axiom
  | axB  : form -> axiom
  | axK4 : form -> axiom
  | axD  : form -> axiom
  | axK5 : form -> axiom.
\end{minted}
Each constructor of type \mintinline{coq}{axiom} represents an axiom and
its type stores the subformulas which compose the axiom. The type has
17 constructors: the first 10 consists encode axioms for propositional logics
and the last 7 are for different formulas taken as axioms by systems of modal logics.
Axioms are built using the function \mintinline{coq}{instantiate}, which
receives an axiom and builds its corresponding formula by pattern matching as
follows:
\begin{minted}{coq}
Definition instantiate (a: axiom): form :=
  match a with
  | ax1    p   q
     => [! p -> (q -> p) !]
  | ax2    p   q   r
     => [! (p -> (q -> r)) -> ((p -> q) -> (p -> r)) !]
  | ax3    p   q
     => [! (~ q -> ~ p) -> (p -> q) !]
  | axK    p   q
     => [! [] (p -> q) -> ([] p -> [] q) !]
  | axPos  p   q
     => [! <> (p \/ q) -> (<> p \/ <> q) !]
  (* some code omitted *)
  end.
\end{minted}
Notation \mintinline{coq}{[! p !]} is used to write modal logic formulas using a
syntax closer to mathematical practice using Coq. This is necessary to avoid
parsing ambiguities, since some symbols are also used by Coq's syntax.
Modalities $\Box$ and $\Diamond$ are denoted by \mintinline{coq}{[]}
and \mintinline{coq}{<>}, respectively.

@@

\paragraph{Systems of modal logics} In order to use the deductive system to
check the provability of formulas, we need to specify the axioms that are
included in the system. Our approach is to define predicates
which denote each system. The simplest set of axioms is usually refereed as
\textbf{K} and it includes the axiom $\Box(A \to B) \to (\Box A \to \Box B)$
and all the axioms for propositional logic. We represent system \textbf{K}
by the following inductive type:
\begin{minted}{coq}
Inductive K: axiom -> Prop :=
  | K_ax1: forall p q, K (ax1 p q)
  | K_ax2: forall p q r, K (ax2 p q r)
  (* some code omitted *)
  | K_axK: forall p q, K (axK p q)
\end{minted}
Type \mintinline{coq}{K} is a predicate over axioms: it has one constructor for
each of the valid axioms which form the system. Given a value
of \mintinline{coq}{axiom}, a value of type \mintinline{coq}{K a} denotes a
proof that a given axiom is part of the system \textbf{K}. This is the main
trick used by the constructor \mintinline{coq}{Ax}, of type
\mintinline{coq}{deduction}, to generically represent axioms for
different systems of modal logics uniformly. In order to understand how it
works, let us review its type (below, remember that \mintinline{coq}{A: axiom -> Prop}
denotes a system of axioms for modal logics):
\begin{minted}{coq}
Ax: forall (t: theory)
           (a: axiom)
           (f: form),
           A a -> instantiate a = f -> deduction A t f
\end{minted}
Note that the first parameter of constructor \mintinline{coq}{Ax} is a value
of type \mintinline{coq}{A a}, which is only inhabited if the predicate
\mintinline{coq}{A} holds for the axiom \mintinline{coq}{a}. Through this
evidence, is possible to create the corresponding axiom formula instance by
using function \mintinline{coq}{instantiate}. In this way, we are able to
represent all systems of axioms for modal logics using a single constructor
for its axioms.

@@

The application of a multimodal separation logic for the semantics of low-level
was the subject of Dockins et. al. work~\cite{Dockins08}. In that work, authors
combine several modal operators that do not need to satisfy the same frame
conditions and seperation logics for reasoning about non-trivial subsets of
low-level programming languages. Authors developed their formalization through a
shallow embedding in Coq proof assistant and provide two case studies: one is
the verification of a simple low-level virtual machine, and the other is the
subset of Sequential Cminor.

>>>>

\paragraph{Syntax and semantics of modal logics}
To implement our library, we use a \emph{deep embedding} of
formulas, i.e. we represent formulas as the following Coq inductive datatype.
\begin{minted}{coq}
Inductive formula: Set :=
| Lit    : nat -> formula
| Neg    : formula -> formula
| Box    : formula -> formula
| Dia    : formula -> formula
| And    : formula -> formula -> formula
| Or     : formula -> formula -> formula
| Implies: formula -> formula -> formula.
\end{minted}
Most constructors of the type have an immediate meaning, since they represent
connectives of propositional logics. The constructor \mintinline{coq}{Lit}
is used to construct propositional variables; constructor \mintinline{coq}{Box}
denotes the modal operator $\Box$ and \mintinline{coq}{Dia}, the operator
$\Diamond$. Furthermore, the type \mintinline{coq}{theory} is defined as a list
of formulas.

\begin{minted}{coq}
Definition theory := list formula.
\end{minted}

Before defining the Kripke semantics, we need to represent the model used to
interpret a modal logic formula. We encode a frame by a record formed by
a type \mintinline{coq}{W: Set} to represent the set of worlds and an
accessibility relation denoted by a type \mintinline{coq}{R: W -> W -> Prop}.
\begin{minted}{coq}
Record Frame: Type := {
  W: Set;
  R: W -> W -> Prop;
}.
\end{minted}
A model is then denoted by another record, formed by a frame and a labeling function,
which assign the truth value of a variable at a given world.
\begin{minted}{coq}
Record Model: Type := {
  F: Frame; 
  v: nat -> (W F) -> Prop; 
}.
\end{minted}
Note that the type of the labeling function \mintinline{coq}{v} ensures that it
refers to a world from the set of possible worlds in the frame
\mintinline{coq}{F}\footnote{We recall that \mintinline{coq}{W F} is Coq's syntax for
accessing field \mintinline{coq}{W} from the record \mintinline{coq}{F}.}. The encoding of the Kripke semantics is done by a Coq function
that implements the relation of Figure~\ref{fig:kripke}. We omit the code
for propositional logics connectives for brevity. The interesting parts are for
variables (constructor \mintinline{coq}{Lit}) which uses the labeling function
\mintinline{coq}{v} from the model \mintinline{coq}{M}. The meaning of
modalities $\Box$ and $\Diamond$ are represented by a quantifying over worlds
using the accessibility relation $R$. 
\begin{minted}{coq}
Fixpoint formula_valuation (M: Model) (w: W (F M))
                           (phi: formula): Prop :=
  match phi with
  | Lit     x
     => v M x w (* should be inhabited, i.e., true *)
  | Box     psi
     => forall w': W (F M), R (F M) w w' ->
                            formula_valuation M w' psi
  | Dia     psi
     => exists w': W (F M), R (F M) w w' /\
                            formula_valuation M w' psi
  (* some code omitted *)
  end.
\end{minted}
In order to ease the task of writing formulas and its semantics properties, we
define Coq notations that mimics the standard way of writing them.

\paragraph{The Hilbert deductive system}
Basically, a Hilbert-style deductive system allows the use of four
possibilities in a proof: 1) the use of a hypothesis; 2) the use of an axiom;
3) the use of the modus ponens rule or 4) the use of the generalization (necessity)
rule. The inductive type \mintinline{coq}{deduction} encodes these possibilities
as its constructors. The \mintinline{coq}{deduction} type is parameterized by
the set of axioms that form a system for some modal logic. In this way, our definition can be
used by several systems alike.
\begin{minted}{coq}
Inductive deduction (A: axiom -> Prop):
                          theory -> formula -> Prop :=
  (* Premise. *)
  | Prem: forall (t: theory)
                 (f: formula)
                 (i: nat),
          (nth_error t i = Some f) -> deduction A t f
  (* Axiom. *)
  | Ax: forall (t: theory)
               (a: axiom)
               (f: formula),
        A a -> instantiate a = f -> deduction A t f
  (* Modus Ponens. *)
  | Mp: forall (t: theory)
               (f g: formula)
               (d1: deduction A t (f -> g))
               (d2: deduction A t f),
        deduction A t g
  (* Generalization. *)
  | Nec: forall (t: theory)
                (f: formula)
                (d1: deduction A t f),
         deduction A t (Box f).
\end{minted}
The constructors \mintinline{coq}{Prem}, \mintinline{coq}{Mp} and \mintinline{coq}{Nec}
represent rules for hypothesis, modus ponens and generalization, respectively.
Before discussing the \mintinline{coq}{Ax}, we need to discuss how axioms are
represented in our library. The inductive type \mintinline{coq}{axiom} specifies
the different axioms supported by our library.
\begin{minted}{coq}
Inductive axiom: Set :=
  | ax1  : formula -> formula -> axiom
  | ax2  : formula -> formula -> formula  -> axiom
  (* some code omitted *)
  | axK  : formula -> formula -> axiom
  | axPos: formula -> formula -> axiom
  | axT  : formula -> axiom
  | axB  : formula -> axiom
  | axK4 : formula -> axiom
  | axD  : formula -> axiom
  | axK5 : formula -> axiom.
\end{minted}
Each constructor of type \mintinline{coq}{axiom} represents an axiom and
its type stores the subformulas which compose the axiom. The type has
17 constructors: the first 10 consists encode axioms for propositional logics
and the last 7 are for different formulas taken as axioms by systems of modal logics.
Axioms are built using the function \mintinline{coq}{instantiate}, which
receives an axiom and builds its corresponding formula by pattern matching as
follows:
\begin{minted}{coq}
Definition instantiate (a: axiom): formula :=
  match a with
  | ax1    p   q
     => [! p -> (q -> p) !]
  | ax2    p   q   r
     => [! (p -> (q -> r)) -> ((p -> q) -> (p -> r)) !]
  | ax3    p   q
     => [! (~ q -> ~ p) -> (p -> q) !]
  | axK    p   q
     => [! [] (p -> q) -> ([] p -> [] q) !]
  | axPos  p   q
     => [! <> (p \/ q) -> (<> p \/ <> q) !]
  (* some code omitted *)
  end.
\end{minted}
Notation \mintinline{coq}{[! p !]} is used to write modal logic formulas using a
syntax closer to mathematical practice using Coq. This is necessary to avoid
parsing ambiguities, since some symbols are also used by Coq's syntax.
Modalities $\Box$ and $\Diamond$ are denoted by \mintinline{coq}{[]}
and \mintinline{coq}{<>}, respectively.

@@

\paragraph{Systems of modal logics} In order to use the deductive system to
check the provability of formulas, we need to specify the axioms that are
included in the system. Our approach is to define predicates
which denote each system. The simplest set of axioms is usually refereed as
\textbf{K} and it includes the axiom $\Box(A \to B) \to (\Box A \to \Box B)$
and all the axioms for propositional logic. We represent system \textbf{K}
by the following inductive type:
\begin{minted}{coq}
Inductive K: axiom -> Prop :=
  | K_ax1: forall p q, K (ax1 p q)
  | K_ax2: forall p q r, K (ax2 p q r)
  (* some code omitted *)
  | K_axK: forall p q, K (axK p q)
\end{minted}
Type \mintinline{coq}{K} is a predicate over axioms: it has one constructor for
each of the valid axioms which form the system. Given a value
of \mintinline{coq}{axiom}, a value of type \mintinline{coq}{K a} denotes a
proof that a given axiom is part of the system \textbf{K}. This is the main
trick used by the constructor \mintinline{coq}{Ax}, of type
\mintinline{coq}{deduction}, to generically represent axioms for
different systems of modal logics uniformly. In order to understand how it
works, let us review its type (below, remember that \mintinline{coq}{A: axiom -> Prop}
denotes a system of axioms for modal logics):
\begin{minted}{coq}
Ax: forall (t: theory)
           (a: axiom)
           (f: formula),
           A a -> instantiate a = f -> deduction A t f
\end{minted}
Note that the first parameter of constructor \mintinline{coq}{Ax} is a value
of type \mintinline{coq}{A a}, which is only inhabited if the predicate
\mintinline{coq}{A} holds for the axiom \mintinline{coq}{a}. Through this
evidence, is possible to create the corresponding axiom formula instance by
using function \mintinline{coq}{instantiate}. In this way, we are able to
represent all systems of axioms for modal logics using a single constructor
for its axioms.

@@

The application of a multimodal separation logic for the semantics of low-level
was the subject of Dockins et. al. work~\cite{Dockins08}. In that work, authors
combine several modal operators that do not need to satisfy the same frame
conditions and seperation logics for reasoning about non-trivial subsets of
low-level programming languages. Authors developed their formalization through a
shallow embedding in Coq proof assistant and provide two case studies:
the verification of a low-level virtual machine and a C-like language.
